Multiplication, division and factorization\\of p-adic polynomials

### 2015-12-23 polynomials/muldivfact.tex
Definition of jagged and Newton precision

\renewcommand{\prec}{\text{\rm prec}}

@@

>>>>

\renewcommand{\prec}{\text{\rm prec}}
\newcommand{\NF}{\text{\rm NF}}

@@

\noindent
{\bf Notations.}
Throughout this paper, we fix a complete discrete valuation field $K$; 
we denote by $\val : K \to \Z \cup \{+\infty\}$ the valuation on it and 
by $W$ its ring of integers (\emph{i.e.} the set of elements with 
nonnegative valuation). We assume that $\val$ is normalized so that it
is surjective and denote by $\pi$ a uniformizer of $K$, that is an 
element of valuation $1$. Denoting by $S \subset W$ a fixed set of 
representatives of the classes modulo $\pi$ and assuming $0 \in S$, 
one can prove that each element in $x \in K$ can be represented 
uniquely as a convergent series:
\begin{equation}
\label{eq:CDVFseries}
x = \sum_{i=\val(x)}^{+\infty} a_i \pi^i
\quad \text{with} \quad a_i \in S.
\end{equation}
The two most important examples are the field of $p$-adic numbers $K = 
\Qp$ and the field of Laurent series $K = k((t))$ over a field $k$. The 
valuation on them are the $p$-adic valuation and the usual valuation of 
a Laurent series respectively. Their ring of integers are therefore 
$\Zp$ and $k[[t]]$ respectively. A distinguished uniformizer is $p$ 
and $t$ whereas a possible set $S$ is $\{0, \ldots, p-1\}$ and $k$
respectively.
The reader who is not familiar with complete discrete valuation fields
may assume (without sacrifying too much to the generality) that $K$ is
one of the two aforementioned examples.

In what follows, the notation $K[X]$ refers to the ring of univariate 
polynomials with coefficients in $K$. The subspace of polynomials of 
degree at most $n$ (resp. exactly $n$) is denoted by $K_{\leq n}[X]$ 
(resp. $K_{=n}[X]$).

\section{Precision data}

Elements in $K$ (and \emph{a fortiori} in $K[X]$) carry an infinite 
amount of information. They thus cannot be stored entirely in the
memory of a computer and have to be truncated. Elements of $K$ are
usually represented by truncating Eq.\eqref{eq:CDVFseries} as
follows:
\begin{equation}
\label{eq:CDVFseriesO}
x = \sum_{i=v}^{N-1} a_i \pi^i + O(\pi^N)
\end{equation}
where $N$ is an integer called the \emph{absolute precision} and 
the notation $O(\pi^N)$ means that the coefficients $a_i$ for $i
\geq N$ are discarded. If $N > v$ and $a_v \neq 0$, the integer $v$ 
is the valuation of $x$ and the difference $N-v$ is called the
\emph{relative precision}.
Alternatively, one may thing that the writing~\eqref{eq:CDVFseriesO}
represents a subset of $K$ which consists of all elements in $K$ for
which the $a_i$'s in the range $[v,N-1]$ are those specified. From the
metric point of view, this is a ball (centered at any point inside it).

It worths be noting that tracking precision using this representation is 
rather easy. For example, if $x$ and $y$ are known with absolute (resp. 
relative) precision $N_x$ and $N_y$ respectively, one can compute the 
sum $x+y$ (resp. the product $xy$) at absolute (resp. relative) 
precision $\min(N_x,N_y)$. Computations with $p$-adic and Laurent
series are often handle this way on symbolic computation softwares.

\subsection{Precision for polynomials}

The situation is much more subtle when we are working with a collection 
of elements of $K$ (\emph{e.g.} a polynomial) and not just a single one.
Indeed, several precision data may be considered and, as we shall see
later, each of them has is own interest. Below we detail three of them
for the special case of polynomials.

\medskip

\noindent
{\bf Jagged precision.}
The most naive precision we may consider is the so-called \emph{jagged 
precision}. It consists in attaching an individual precision (of the
shape $O(\pi^{N_i})$) to each coefficient separatedly.

\medskip

\noindent
{\bf Newton precision.} \todo{Write a one-line introduction.}

\begin{deftn}
A \emph{Newton function of degree $n$} is a convex function 
$f : [0,n] \to \R$ which is piecewise affine and whose epigraph
$\Epi(f)$ have integral extremal points.
\end{deftn}

\begin{rem}
The datum of $f$ is equivalent to that of $\Epi(f)$ and they can
easily represented and manipulated on a computer.
\end{rem}

We recall that one can attach a Newton function to each polynomial.
If $P(X) = \sum_{i=0}^n a_n X^n \in K_n[X]$, we define its Newton
polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
($1 \leq i \leq n$) together with the point at infinity $(0,+\infty)$
and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
function whose epigraph is $\NP(P)$. It is well known \todo{(Give a
reference?)} that:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NP(P+Q) & \subset & \text{Conv}\big(\NP(P) \cup \NP(Q)\big) \smallskip \\
\NP(PQ) & = & \NP(P) + \NP(Q)
\end{array}$$
where $\text{Conv}$ denotes the convex hull and the plus sign stands 
for the Minkowski sum. This translates to:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NF(P+Q) \geq \NF(P) \oplus \NF(Q) \smallskip \\
\NF(PQ) = \NF(P) \odot \NF(Q)
\end{array}$$
where the operations $\oplus$ and $\odot$ are defined accordingly.
It turns out that there exist efficient algorithms for computing these 
operations.

In a similar fashion, Newton functions can be used to model precision: 
given a Newton function $f$ of degree $n$, we agree that a polynomial 
of degree at most $n$ is given at precision $O(f)$ when, for all $i$,
its $i$-th coefficient is given at precision $O\big(\pi^{\lceil f(i)
\rceil}\big)$ (where $\lceil \cdot \rceil$ is the ceiling function).
In the sequel, we shall write
$O(f) = \sum_{i=0}^n O\big(\pi^{\lceil f(i) \rceil}\big) \cdot X^i$
and use the notation $\sum_{i=0}^n a_i X^i + O(f)$
to refer to a polynomial given at precision $O(f)$.

We observe in particular that Newton precision data are particular
instances of jagged precision. The main interest of using them is the
efficiency.

\medskip

\noindent
{\bf Lattice precision.}
